Distributional Trees for Circular Data

### 2019-04-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
first draft of extended abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2516 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 82bb06a3b27bdc2fb6e10702f8102edf95e55ef2

%\section{Motivation}
%Circular response variables occur in a variety of applications and subject areas. 
%For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
%animal orientation or gene-structure analysis are often subject of examination in biology, 
%and wind direction is one of the most important weather variables in meteorology.
%Representing such a two-dimensional direction by an angle in a unity circle requires 
%the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
%rotation. However, these settings can vary but should not influence statistical 
%inference from a fitted model.
%For that reason circular data demand for purpose-built statistical methods. 
%A commonly chosen distribution for modeling circular data is the von Mises distribution.
%Estimating its two distribution parameters by generalized additive models using a set of 
%regressor variables is one possible approach that can be successful in various settings.
%As an alternative we propose the application of distributional trees 
%which are capable of detecting also non-additive effects and provide easily interpretable 
%models allowing for a wide range of inference procedures.
%Moreover, covariates and their possible interactions are selected automatically
%such that no expert knowledge is required to specify the model in advance.

@@

Distributional trees~(Schlosser et.~al, 2018) fuse distributional regression modeling with
regression trees based on the unbiased recursive partitioning algorithms MOB~(Zeileis et.~al, 2008)
or CTree~(Hothorn et.~al, 2006). The basic idea is to partition the covariate space recursively into 
subgroups such that an (approximately) homogeneous distributional model can be fitted to the response 
in each resulting subgroup.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test (MOB) or a permutation test (CTree).
In each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood (MOB) or a two-sample test statistic (CTree)
over all possible partitions.

@@

\begin{figure}[p!]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{schlosser-circtree_plot.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
all lagged by one hour.}
\label{schlosser:fig_tree} \end{figure}

@@

A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model
(see Schlosser et.~al~(2018) for a general discussion of distributional forests).

Another possible improvement for obtaining more parsimonious and more stable trees is to
consider splitting circular covariates into two circle segments by searching two
split points simultaneously (rather than sequentially at different depths as
for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).

@@

@@

%\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
%     Classification and Regression Trees.
%     {\it Wadsworth}.

@@

@@

\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     arXiv:1804.02921, {\it arXiv.org E-Print Archive}.

>>>>

\section{Motivation}
\textit{FIXME: should we include this section 'Motivation'?}

Circular response variables occur in a variety of applications and subject areas. 
For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
animal orientation or gene-structure analysis are often subject of examination in biology, 
and wind direction is one of the most important weather variables in meteorology.
Representing such a two-dimensional direction by an angle in a unity circle requires 
the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
rotation. However, these settings can vary but should not influence statistical 
inference from a fitted model.
For that reason circular data demand for purpose-built statistical methods. 
%A commonly chosen distribution for modeling circular data is the von Mises distribution.
%Estimating its two distribution parameters by generalized additive models using a set of 
%regressor variables is one possible approach that can be successful in various settings.
%As an alternative we propose the application of distributional trees 
%which are capable of detecting also non-additive effects and provide easily interpretable 
%models allowing for a wide range of inference procedures.
%Moreover, covariates and their possible interactions are selected automatically
%such that no expert knowledge is required to specify the model in advance.

@@

Distributional trees~(Schlosser et.~al, 2019) fuse distributional regression modeling with
regression trees based on the unbiased recursive partitioning algorithms MOB~(Zeileis et.~al, 2008)
or CTree~(Hothorn et.~al, 2006). The basic idea is to partition the covariate space recursively into 
subgroups such that an (approximately) homogeneous distributional model can be fitted to the response 
in each resulting subgroup.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test (MOB) or a permutation test (CTree).
In each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood (MOB) or a two-sample test statistic (CTree)
over all possible partitions.

@@

%% exchange file name of figure: schlosser-circtree_plot.pdf -> schlosserfig1.pdf
\begin{figure}[p!]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{schlosser-circtree_plot.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west and east of the airport,
all lagged by one hour.}
\label{schlosser:fig_tree} \end{figure}

@@

@@

\subsection{Ensembles/Random Forests}
A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by Breiman~et.~al~(1984) average the
predictions of an ensemble of trees, each built on a subsample 
or bootstrap of the original data. 
A different strategy of combining trees is employed in distributional forests 
(Schlosser~et.~al, 2019) which can also be seen as adaptive local likelihood models. 
For a possibly new observation $x$ weights $w_i(x)$ are obtained by counting the 
number of trees in which it is assigned to the same terminal node as a learning
observation $y_i, i \in \{1,\ldots,n\}$.
The parameters $\mu$ and $\kappa$ can then be estimated specifically for the (new) 
observation $x$ by applying the weighted maximum likelihood estimator
$$
\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
$$
In that way predictions are made mainly based on similar observations rather than the 
complete learning data.

\subsection{Splits in Circular Variables}
Another possible improvement for obtaining more parsimonious and more stable trees is to
consider splitting circular covariates into two circle segments by searching two split 
points simultaneously (rather than sequentially at different depths as for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
In particular, the search for the best separation of the data space into two intervals is based on the optimization problem
$$
\max \sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)
$$
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters for the corresponding subsets of the response as defined by the intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
In case of a linear split variable these intervals can be defined by a single split point. 
By treating a circular variable such as wind direction linearly as done in the presented 
application the considered space~$[0,2\pi)$ is then split by a split point $\nu$ into
$\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing an unnatural separation at $0$. 
Moreover, further splits are often required to obtain properly fitting subintervals of~$[0,2\pi)$.
This can be avoided by searching for two split points $\nu$ and $\tau$ within the maximization
process such that $\mathit{left}=[\nu,\tau]$ and $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$.
The required test statistic to maximally select two split points simultaneously is readily 
available in the \textsf{R} package \texttt{coin} and could be enhanced as proposed by Hothorn 
and Zeileis~(2008). For a score-based approach the general strategy of double maximum tests 
introduced by Bai and Perron~(2003) can be considered. 
Hence, the splitting idea can be extended naturally to a two point-search, however, for 
an unbiased and inference based selection the corresponding tests employing a proper
asymptotic null distribution still need to be implemented.

@@

\item[Bai, J., and Perron, P.] (2003).
     Computation and Analysis of Multiple Structural Change Models.
     {\it Journal of Applied Econometrics}, {\bf 18},
     1\,--\,22.

\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
     Classification and Regression Trees.
     {\it Wadsworth}.

@@

\item[Hothorn, T., and Zeileis, A.] (2008).
     Generalized Maximally Selected Statistics.
     {\it Biometrics}, {\bf 64}, 4,
     1263\,--\,1269.

@@

\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2019).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     arXiv:1804.02921, {\it arXiv.org E-Print Archive}.
